Learning and comparing functional connectomes across subjects

### 2012-12-30 paper.tex
minor edits

\paragraph{FMRI-based function definition}
%
Defining regions directly from the fMRI signal brings many benefits.
First, it can capture subject-specific functional information. Second, it
adapts to the signal at hand and its limitations, such as
EPI-specific warps or vascular and movement artifacts that are isolated
in ICA-like approaches. The simplest approach to define task-specific
regions is to use activation maps derived from standard GLM-based
analysis in a task-driven study (see for instance \cite{poldrack2011}).
Regions are extracted by thresholding the maps, or using balls around the
activation peaks. For resting-state studies, unsupervised multivariate
analysis techniques are necessary. Clustering approaches extract
full-brain parcellations \cite{craddock2011, bellec2010, yeo2011,
thirion2006}, and have been shown to extract well-known functional
structures from rest data \cite{yeo2011}. Alternatively, decompositions
methods, such as ICA \cite{beckmann2004}, can unmixing linear
combinations of multiple effects and separate out multiple spatial maps
capturing functional networks or confounding effects. At high model
order, ICA maps define a functional parcellation \cite{kiviniemi2009}.
Extracting regions from these maps requires additional effort as they can
display fragmented spatial features and structured background noise, but
incorporating sparsity and spatial constraints in the decomposition
techniques leads to contrasted maps that outline many different
structures \cite{varoquaux2012} (see fig.\,\ref{fig:parcellations}).

@@

\caption{
Correlation matrices of rest time-series extracted from the 39 main
regions of the Varoquaux 2011 \cite{varoquaux2011} parcellation with
different choices of confound regressors -- 
\textbf{Left}: regressing out CompCor signals, as well as white matter and
CSF average signals and movement parameters. The insert shows the
connections restricted to the .
-- \textbf{Upper right}: regressing out only movement parameters. -- 
\textbf{Lower right}:
regressing out movement parameters and global signal mean.
No frequency filtering was applied here.
\label{fig:correlation_matrices}
When no confounding brain signals are regressed, all regions are heavily
correlated. Regressing out common signal, in the form of well-identified
confounds or a global mean, teases out the structure.
}
\end{figure}
%---------------------------------------------------------------------------}}}

@@

\paragraph{Signal extraction}
%
Given ROIs, the next step is to extract from them a summary measure. To
study \emph{on-going} activity, \emph{e.g.}\ with rest data, signal
extraction is performed from the EPI time series; however, to study
\emph{evoked} activity with task-driven studies it can be beneficial to
run a first-level analysis, enforcing specificity of the measure
extracted to the task. With slow event-related design, task-specific
functional connectivity can be captured in trial-to-trial fluctuations in
the BOLD response, estimated using a GLM analysis with one regressor per
trial \cite{grillon2012,rissman2004}. This approach, known as beta-series
regression, has been adapted for rapid event-related designs, using
multiple GLMs to optimize deconvolution of each trial \cite{mumford2012}.
With rest datasets, it is important to regress out time series capturing
movement and other sources of structured noise, to separate on-going
activity from confounding signals. In addition to removing movement
parameters estimated during preprocessing, removing linear trends, signal
extracted from the white matter and the CSF \cite{chang2009}, as well as
using the CompCor \cite{behzadi2007} procedure gives more contrasted
correlation matrices that outline functional structures better
(fig.\,\ref{fig:correlation_matrices}). Filtering high frequencies is
often recommended, based on the initial observation that neural-related
signal are observed below 0.1\,Hz \cite{cordes2001,biswal1995}. While
high-pass and low-pass filtering decrease the impact of some confounds,
recent studies have shown that information on neural processes is present
in the full spectrum of frequencies observed
\cite{smith2012,vanoort2012}. Regressing out a good choice of confound
signals is more specific than frequency filter, and in our experience
gives more contrasted correlation matrices\footnote{Note that naive use
of filtering can induce spurious correlations \cite{davey2012}.}.

@@

\caption{
Different inverse-covariance matrices estimates corresponding to
Fig.\,\ref{fig:correlation_matrices} -- \textbf{Left}: group-sparse estimate
using the $\ell_{21}$ estimator \cite{varoquaux2010c}.
The insert shows the
connections restricted to the default mode network. -- \textbf{Upper
right}: non-sparse estimate: inverse of the sample correlation matrix. --
\textbf{Lower right}: sparse estimate using the Graph Lasso
\cite{friedman2008}.
\label{fig:icov_estimators}
}
\end{figure}
%---------------------------------------------------------------------------}}}

\paragraph{Correlation and partial correlations}
%
Functional connectivity between the ROIs can be measured by computing the
correlation matrix of the extracted signals. An important and often
neglected point is that the sample correlation matrix, \emph{i.e.}\ the
correlation matrix obtained by plugging the observed signal in the
correlation matrix formula, is not the population correlation matrix,
\emph{i.e.}\ the correlation matrix of the data-generating process. If the
number of measurements was infinite, the two would coincide, however if
this number is not large compared to the number of connection (that
scales as the square of the number of ROIs), the sample correlation
matrix is a poor estimate of the underlying population correlation
matrix. In other words, the sample correlation matrix captures a lot of
sampling noise, intrinsic randomness that arises in the estimation of
correlations from short time series. Conclusions drawn from the sample
correlation matrix can easily reflect this estimation error.
%
Varoquaux \emph{et al.}~\cite{varoquaux2010c} and Smith \emph{et
al.}~\cite{smith2011} have shown respectively on rest fMRI and on
realistic simulations that a good choice of correlation matrix estimator
could recover the connectivity structure, where the sample correlation
matrix would fail. In general, the choice of a better estimate depends on
the settings and the end goals \cite{varoquaux2012,varoquaux2010b},
however the Ledoit-Wolf shrinkage estimate \cite{ledoit2004} is a simple,
computationally-efficient, and parameter-less alternative that performs
uniformly better than the sample correlation matrix
\cite{varoquaux2012,varoquaux2010c} and should always be preferred.

@@

For the problem of recovering the functional-connectivity
\emph{structure}, \emph{i.e.}\ finding which region is connected to
which, sparse inverse covariance estimators have been found to be
efficient \cite{varoquaux2010c, smith2011}. The intuition for relying on
inverse covariance rather than correlation stems from that fact that
standard correlation (marginal correlation) between two variables $a$ and
$b$ also capture the effects of other variables -- strong correlation of
$a$ and $b$ with a third variable $c$ will induce a correlation between
$a$ and $b$. On the opposite, the inverse covariance\footnote{Covariance
and correlation matrices differ simply by the fact that a covariance
matrix captures the amplitude of a signal, via its variance, while a
correlation matrix is computed on normalized signals.}
matrix captures
partial correlations, removing the effect of other variables
\cite{marrelec2006a}. In the small sample limit, this removal is
challenging from the statistical standpoint, and assumption of sparsity,
\emph{i.e.}\ that only few variables need to be considered at a time, is
important to estimate a good inverse covariance. Various estimation
strategies exist for sparse inverse covariance, and have an impact on the
resulting networks \cite{varoquaux2012,varoquaux2010c}. The
Graph Lasso ($\ell_1$-penalized maximum-likelihood estimator)
\cite{friedman2008} is in general a good approach for structure recovery. In group studies,
the $\ell_{21}$ estimator \cite{varoquaux2010c,honorio2012} is useful to
impose a common sparsity structure across different subjects and achieve
better recovery of this common structure. Simply put, these approaches
are necessary because estimation noise creates a background structure
(see Fig.\,\ref{fig:icov_estimators}); however, unlike in a univariate situation, the
parameters are not independent, and the spurious background connections
degrade the estimation of the actual connections. The sparse estimators
make a compromise between imposing simpler models, \emph{i.e.}\ with less
connections, and fitting well the data. This compromise is set via a
regularization parameter which controls the sparsity of the estimate. A
good rule of thumb to set this parameter is via cross-validation
\cite{varoquaux2010c}.

@@

\paragraph{Network structure extracted}
%
The correlation matrices and inverse-covariance matrices that we extract
contain a lot of information on the functional structure of the brain.
First, the correlation matrix (Fig.\,\ref{fig:correlation_matrices})
shows blocks of coactivating regions that can be interpreted as
large-scale functional networks, such as the default mode network. Note
the split in network is not straightforward. Different ordering of the
nodes will reveal different networks. Indeed, because of the presence of
hubs and interleaved networks the picture in terms of segregated networks
is not sufficient to explain full-brain connectivity
\cite{varoquaux2012}. The connectivity matrices, correlation matrix or
inverse-covariance matrix, can be represented as graphs: nodes connected
by weighted edges (inserts on Fig.\,\ref{fig:correlation_matrices} and
Fig.\,\ref{fig:icov_estimators}). The inverse-covariance matrix,
capturing partial correlations, appears then as extracting a
\emph{backbone} or \emph{core} of the graph. While such structure have
been used as a way to summarize anatomical brain connectivity graphs
\cite{hagmann2008}, here they have a clear-cut meaning with regards to
the BOLD signal: they give the conditional independence structure between
the regions \cite{varoquaux2012}. In other words, regions $a$ and $b$ are
not connected if the signal that they have in common can be explained by
a third region $c$. In this light, the choice of nuisance regressors to
remove confounding common signal is less critical. Indeed, while with
correlation matrices regressing out the global mean has a drastic
(Fig.\,\ref{fig:correlation_matrices} upper right and lower right), on
inverse covariance it only changes very slightly the resulting matrices
(Fig.\,\ref{fig:icov_regressors}). There have been debates on whether to
regress out certain signals, such as the global mean,  as it induces
negative correlations \cite{murphy2009,chang2009,fox2009}, and these may seem
surprising: one network is doing the opposite of the other. However,
correlation between two signals only takes its meaning with the
definition of a baseline. A simple picture to explain anti-correlations
between two regions is the presence of a third region, mediating the
interactions. Using this third region as a baseline would amount to
estimate partial correlations in the whole system. Using 
inverse-covariance matrices or partial correlations to understand brain
connectivity makes the interpretation in terms of interactions between
brain regions easier and more robust to the choice of confounds.

>>>>

\paragraph{FMRI-based function definition}
%
Defining regions directly from the fMRI signal brings many benefits.
First, it can capture subject-specific functional information. Second, it
adapts to the signal at hand and its limitations, such as
EPI-specific warps or vascular and movement artifacts that are isolated
in ICA-like approaches. The simplest approach to define task-specific
regions is to use activation maps derived from standard GLM-based
analysis in a task-driven study (see for instance \cite{poldrack2011}).
Regions are extracted by thresholding the maps, or using balls around the
activation peaks. For resting-state studies, unsupervised multivariate
analysis techniques are necessary. Clustering approaches extract
full-brain parcellations \cite{craddock2011, bellec2010, yeo2011,
thirion2006}, and have been shown to extract well-known functional
structures from rest data. Alternatively, decompositions
methods, such as ICA \cite{beckmann2004}, can unmix linear
combinations of multiple effects and separate out overlapping spatial maps
capturing functional networks or confounding effects. At high model
order, ICA maps define a functional parcellation \cite{kiviniemi2009}.
Extracting regions from these maps requires additional effort as they can
display fragmented spatial features and structured background noise, but
incorporating sparsity and spatial constraints in the decomposition
techniques leads to contrasted maps that outline many different
structures \cite{varoquaux2012} (see fig.\,\ref{fig:parcellations}).

@@

\caption{
Correlation matrices of rest time-series extracted from the 39 main
regions of the Varoquaux 2011 \cite{varoquaux2011} parcellation with
different choices of confound regressors -- 
\textbf{Left}: regressing out CompCor signals, as well as white matter and
CSF average signals and movement parameters. The insert shows the
connections restricted to a few major nodes.
-- \textbf{Upper right}: regressing out only movement parameters. -- 
\textbf{Lower right}:
regressing out movement parameters and global signal mean.
No frequency filtering was applied here.
\label{fig:correlation_matrices}
When no confounding brain signals are regressed, all regions are heavily
correlated. Regressing out common signal, in the form of well-identified
confounds or a global mean, teases out the structure.
}
\end{figure}
%---------------------------------------------------------------------------}}}

@@

\paragraph{Signal extraction}
%
Given ROIs, the next step is to extract from them a summary measure. To
study \emph{on-going} activity, \emph{e.g.}\ with rest data, signal
extraction is performed from the EPI time series; however, to study
\emph{evoked} activity with task-driven studies it can be beneficial to
run a first-level analysis, enforcing specificity of the measure
extracted to the task. With slow event-related designs, task-specific
functional connectivity can be captured in trial-to-trial fluctuations in
the BOLD response, estimated using a GLM analysis with one regressor per
trial \cite{grillon2012,rissman2004}. This approach, known as beta-series
regression, has been adapted for rapid event-related designs, using
multiple GLMs to optimize deconvolution of each trial \cite{mumford2012}.
With rest datasets, it is important to regress out time series capturing
movement and other sources of structured noise, to separate on-going
activity from confounding signals. In addition to removing movement
parameters estimated during preprocessing, removing linear trends, signal
extracted from the white matter and the CSF \cite{chang2009}, as well as
using the CompCor \cite{behzadi2007} procedure gives more contrasted
correlation matrices that outline functional structures better
(fig.\,\ref{fig:correlation_matrices}). Filtering high frequencies is
often recommended, based on the initial observation that neural-related
signals are observed below 0.1\,Hz \cite{cordes2001,biswal1995}. While
high-pass and low-pass filtering decrease the impact of some confounds,
recent studies have shown that information on neural processes is present
in the full spectrum of frequencies observed
\cite{smith2012,vanoort2012}. Regressing out a good choice of confound
signals is more specific than frequency filter, and in our experience
gives more contrasted correlation matrices\footnote{Note that naive use
of filtering can induce spurious correlations \cite{davey2012}.}.

@@

\caption{
Different inverse-covariance matrices estimates corresponding to
Fig.\,\ref{fig:correlation_matrices} -- \textbf{Left}: group-sparse estimate
using the $\ell_{21}$ estimator \cite{varoquaux2010c}.
The insert shows the
connections restricted to a few major nodes. -- \textbf{Upper
right}: non-sparse estimate: inverse of the sample correlation matrix. --
\textbf{Lower right}: sparse estimate using the Graph Lasso
\cite{friedman2008}.
\label{fig:icov_estimators}
}
\end{figure}
%---------------------------------------------------------------------------}}}

\paragraph{Correlation and partial correlations}
%
Functional connectivity between the ROIs can be measured by computing the
correlation matrix of the extracted signals. An important and often
neglected point is that the sample correlation matrix, \emph{i.e.}\ the
correlation matrix obtained by plugging the observed signal in the
correlation matrix formula, is not the population correlation matrix,
\emph{i.e.}\ the correlation matrix of the data-generating process. If the
number of measurements was infinite, the two would coincide, however if
this number is not large compared to the number of connections (that
scales as the square of the number of ROIs), the sample correlation
matrix is a poor estimate of the underlying population correlation
matrix. In other words, the sample correlation matrix captures a lot of
sampling noise, intrinsic randomness that arises in the estimation of
correlations from short time series. Conclusions drawn from the sample
correlation matrix can easily reflect this estimation error.
%
Varoquaux \emph{et al.}~\cite{varoquaux2010c} and Smith \emph{et
al.}~\cite{smith2011} have shown respectively on rest fMRI and on
realistic simulations that a good choice of correlation matrix estimator
could recover the connectivity structure, where the sample correlation
matrix would fail. In general, the choice of a better estimate depends on
the settings and the end goals \cite{varoquaux2012,varoquaux2010b},
however the Ledoit-Wolf shrinkage estimate \cite{ledoit2004} is a simple,
computationally-efficient, and parameter-less alternative that performs
uniformly better than the sample correlation matrix
\cite{varoquaux2012,varoquaux2010c} and should always be preferred.

@@

For the problem of recovering the functional-connectivity
\emph{structure}, \emph{i.e.}\ finding which region is connected to
which, sparse inverse covariance estimators have been found to be
efficient \cite{varoquaux2010c, smith2011}. The intuition for relying on
inverse covariance rather than correlation stems from that fact that
standard correlation (marginal correlation) between two variables $a$ and
$b$ also capture the effects of other variables: strong correlation of
$a$ and $b$ with a third variable $c$ will induce a correlation between
$a$ and $b$. On the opposite, the inverse covariance\footnote{Covariance
and correlation matrices differ simply by the fact that a covariance
matrix captures the amplitude of a signal, via its variance, while a
correlation matrix is computed on normalized signals.}
matrix captures
partial correlations, removing the effect of other variables
\cite{marrelec2006a}. In the small sample limit, this removal is
challenging from the statistical standpoint, and assumption of sparsity,
\emph{i.e.}\ that only few variables need to be considered at a time, is
important to estimate a good inverse covariance. Various estimation
strategies exist for sparse inverse covariance, and have an impact on the
resulting networks \cite{varoquaux2012,varoquaux2010c}. The
Graph Lasso ($\ell_1$-penalized maximum-likelihood estimator)
\cite{friedman2008} is in general a good approach for structure recovery. In group studies,
the $\ell_{21}$ estimator \cite{varoquaux2010c,honorio2012} is useful to
impose a common sparsity structure across different subjects and achieve
better recovery of this common structure. Simply put, these approaches
are necessary because estimation noise creates a background structure
(see Fig.\,\ref{fig:icov_estimators}); however, unlike in a univariate situation, the
parameters are not independent, and the spurious background connections
degrade the estimation of the actual connections. The sparse estimators
make a compromise between imposing simpler models, \emph{i.e.}\ with less
connections, and fitting well the data. This compromise is set via a
regularization parameter which controls the sparsity of the estimate. A
good rule of thumb to choose this parameter is via cross-validation
\cite{varoquaux2010c}.

@@

\paragraph{Network structure extracted}
%
The correlation matrices and inverse-covariance matrices that we extract
contain a lot of information on the functional structure of the brain.
First, the correlation matrix (Fig.\,\ref{fig:correlation_matrices})
shows blocks of coactivating regions that can be interpreted as
large-scale functional networks, such as the default mode network. Note
that the split in networks is not straightforward. Different ordering of the
nodes will reveal different networks. Indeed, because of the presence of
hubs and interleaved networks, the picture in terms of segregated networks
is not sufficient to explain full-brain connectivity
\cite{varoquaux2012}. The connectivity matrices, correlation matrix or
inverse-covariance matrix, can be represented as graphs: nodes connected
by weighted edges (inserts on Fig.\,\ref{fig:correlation_matrices} and
Fig.\,\ref{fig:icov_estimators}). The inverse-covariance matrix,
capturing partial correlations, appears then as extracting a
\emph{backbone} or \emph{core} of the graph. While such structure have
been used as a way to summarize anatomical brain connectivity graphs
\cite{hagmann2008}, here they have a clear-cut meaning with regards to
the BOLD signal: they give the conditional independence structure between
the regions \cite{varoquaux2012}. In other words, regions $a$ and $b$ are
not connected if the signal that they have in common can be explained by
a third region $c$. In this light, the choice of nuisance regressors to
remove confounding common signal is less critical with partial
correlations than with correlations. Indeed, while with
correlation matrices regressing out the global mean has a drastic
(Fig.\,\ref{fig:correlation_matrices} upper right and lower right), on
inverse covariance it only changes very slightly the resulting matrices
(Fig.\,\ref{fig:icov_regressors}). There have been debates on whether to
regress out certain signals, such as the global mean,  as it induces
negative correlations \cite{murphy2009,chang2009,fox2009}, and these may seem
surprising: one network is doing the opposite of the other. However,
correlation between two signals only takes its meaning with the
definition of a baseline. A simple picture to explain anti-correlations
between two regions is the presence of a third region, mediating the
interactions. Using this third region as a baseline would amount to
estimate partial correlations in the whole system. Using 
inverse-covariance matrices or partial correlations to understand brain
connectivity makes the interpretation in terms of interactions between
brain regions easier and more robust to the choice of confounds.
